Parameter Recovery Validation

### 2025-06-15 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): simplify temporal prior
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving critical dimensional analysis inconsistencies identified in the piecewise activation framework. The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters. Additionally, this study implements a **process duration with stick-breaking** temporal framework that eliminates scaling symmetry and achieves parameter identifiability while preserving all analytical and computational advantages.

@@

### Process Duration with Stick-Breaking Framework

This study implements a **process duration with stick-breaking** approach that eliminates the scaling symmetry present in hierarchical temporal scaling while maintaining all analytical and biological advantages. The key innovation is the use of **fixed anchor points** that break scaling symmetry and achieve parameter identifiability.

#### **Elimination of Scaling Symmetry**

The previous hierarchical scaling approach suffered from fundamental **scaling symmetry** that prevented parameter identifiability:

$$\mathcal{L}(T_M^*, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) = \mathcal{L}(\lambda T_M^*, \{\lambda t^*_j\}, \{\lambda t^*_{0,\text{on}i}\}, \{\lambda \delta^*_i\})$$

for any $\lambda > 0$. This symmetry creates non-identifiable parameters and leads to catastrophic validation failures.

#### **Fixed Anchor Point Solution**

The stick-breaking approach eliminates scaling symmetry through **fixed anchor points**:

\begin{align}
t^*_1 &= 0 \quad \text{(process start, fixed)} \\
t^*_N &= T_{\max} \quad \text{(process end, sampled)} \\
t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{\text{on}}, \sigma_{\text{on}}^2) \quad \text{(independent gene timing)} \\
\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(independent gene duration)}
\end{align}

**Theorem (Elimination of Scaling Symmetry):**

For any scaling factor $\lambda \neq 1$, the likelihood functions are not equal:
$$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$

**Proof:**
1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
2. **Process duration independence**: Gene parameters $t^*_{0,\text{on}i}, \delta^*_i$ are sampled independently of $T_{\max}$ and do not scale
3. **Asymmetric scaling**: Scaling $T_{\max} \rightarrow \lambda T_{\max}$ and $t^*_j \rightarrow \lambda t^*_j$ changes the relative positioning of gene events within the process timeline
4. **Likelihood dependence**: The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$, which change under asymmetric scaling

Therefore, the likelihood is not invariant under parameter scaling, eliminating the scaling symmetry. □

@@

This framework enables representation of diverse gene expression patterns through specific parameter combinations. Each pattern is defined by quantitative parameter ranges applied to **absolute temporal parameters** within the process duration framework.

**Parameter Framework**: The classification uses absolute temporal parameters that are independent of the process duration $T_{\max}$, enabling robust pattern identification across different experimental timescales.

@@

- $t^*_{0,\text{on}i} < 0.0$ (activation onset before process start)
- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)

**Biological interpretation**: These genes underwent activation before the biological process under study began ($t^* = 0$). During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels. The negative onset time indicates that activation occurred before the process boundary.

@@

**Characteristics**: Genes that show complete activation-decay cycles within the process duration, exhibiting brief, pulse-like activation followed by return toward baseline.

@@

- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
- $t^*_{0,\text{on}i} < 2.5$ (activation early enough to complete cycle within process)
- $\delta^*_i < 2.0$ (short activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the process duration. The constraints ensure that both the activation and subsequent decay phases are captured within the temporal boundaries.

@@

**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the process duration, showing net increase over the process timeline.

@@

- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
- $t^*_{0,\text{on}i} < 1.5$ (early activation onset within process)
- $\delta^*_i > 2.5$ (long activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

**Biological interpretation**: These genes activate early in the process and maintain high expression levels throughout most of the process duration. The long activation duration means that decay back to baseline is either minimal or occurs beyond the process endpoint.

@@

@@

The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when the biological process begins. This extension is mathematically valid and biologically realistic:

@@

1. **Phase determination remains consistent**: For any process time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$

@@

3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the process start ($t^* = 0$)

@@

- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no cells will be in Phase 1 during the process
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the process duration

### Constraint Feasibility Analysis and Optimization

The pattern constraints above were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability. The analysis now includes both LogNormal and Normal distributions to handle the extended temporal framework.

@@

# Prior hyperparameters (updated for absolute temporal parameterization)
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},              # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 2.5, 'scale': 1.5},           # Normal(2.5, 1.5²), absolute onset (allows negatives)
    'delta_star': {'loc': 0.0, 'scale': 0.5},          # log(1.0), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for absolute parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before process start
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),        # Activation within process duration
        ('t_on_star', '<', 2.5),        # Early enough to complete cycle
        ('delta_star', '<', 2.0),       # Short pulse duration
        ('R_on', '>', 2.0),             # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),        # Activation within process duration
        ('t_on_star', '<', 1.5),        # Early activation onset
        ('delta_star', '>', 2.5),       # Long activation duration
        ('R_on', '>', 2.0),             # Strong fold change
    ]
}

@@

The probabilistic graphical model implements the process duration with stick-breaking framework while adding gene-specific piecewise activation parameters. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of absolute temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

# hyperparameters for process duration
pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)

# hyperparameters for lambda (entering from right edge of cell plate)
pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 7.5, 3.5, fixed=True)
pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 7.5, 2.5, fixed=True)

@@

# process duration parameter (replaces hierarchical time scaling)
pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)

# stick-breaking cell times and cell-specific parameters
pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
pgm.add_node("lambda_j", r"$\lambda_j$", 6.5, 3)

@@

# process duration edges
pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
pgm.add_edge("beta_T", "T_max", plot_params=edge_params)

# stick-breaking time edges
pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)

@@

The model combines the process duration with stick-breaking framework with piecewise constant transcription rates and steady-state initial conditions to achieve parameter identifiability while enabling rich gene expression dynamics.

@@

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **absolute temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- $T_{\max}$ represents the total process duration with fixed anchor points at $t^*_1 = 0$ and $t^*_N = T_{\max}$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- Gene temporal parameters are sampled independently of $T_{\max}$, eliminating scaling symmetry

**Detailed priors for process duration with stick-breaking framework:**

The priors are designed to encode biologically reasonable parameter ranges while achieving parameter identifiability through fixed anchor points and independent gene timing:

**Process duration:**
\begin{align}
    T_{\max} &\sim \text{Gamma}(5.0, 1.0) \quad \text{(total process duration)}
\end{align}

@@

**Absolute temporal gene-specific parameters (independent of $T_{\max}$):**
\begin{align}
    t^*_{0,\text{on}i} &\sim \text{Normal}(2.5, 1.5^2) \quad \text{(absolute activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(0.0, 0.5^2) \quad \text{(absolute activation duration)}
\end{align}

**Stick-breaking cell times:**
\begin{align}
    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
    t^*_N &= T_{\max} \quad \text{(process end)} \\
    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
\end{align}

**Pattern Classification with Absolute Parameters:**

Using the expected process duration $\mathbb{E}[T_{\max}] = 5.0$, the pattern constraints become:

- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$ (gene activated before process begins)
- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$ and $R_{\text{on},i} > 2.0$ (complete cycle within process)
- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$ and $R_{\text{on},i} > 2.0$ (long-duration activation)

These absolute thresholds preserve the same biological pattern meanings as the original relative formulation.

**Biological interpretation of prior choices:**

**Process duration:**

- **$T_{\max} \sim \text{Gamma}(5.0, 1.0)$**: Centers process duration at 5 dimensionless time units, providing sufficient range for exponential convergence while maintaining computational efficiency. The process has clear start ($t^* = 0$) and end ($t^* = T_{\max}$) boundaries.

@@

**Absolute temporal parameters:**

- **$t^*_{0,\text{on}i} \sim \text{Normal}(2.5, 1.5^2)$**: Centers activation onset at the middle of the expected process duration, with 95% between approximately -0.4 and +5.4. **Negative values** enable pre-activation scenarios where genes were already activated before the process began.

- **$\delta^*_i \sim \text{LogNormal}(0.0, 0.5^2)$**: Centers activation duration at 1 dimensionless time unit, with 95% between ~0.37 and ~2.7. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): ~4.8% probability under the prior
- **Transient** ($0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles
- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation

**Parameter independence benefits:**

- **Identifiability**: Gene timing parameters are independent of process duration, eliminating scaling symmetry
- **Biological realism**: Process boundaries ($0$ to $T_{\max}$) provide natural temporal reference frame
- **Cross-scale applicability**: Same parameter values work across different experimental timescales
- **Separation of concerns**: Process duration ($T_{\max}$) vs gene-specific timing (onset/duration) vs kinetics (degradation/fold-change)

**Cross-System Biological Interpretation:**

The same $T_{\max} = 5.0$ represents different real-world durations depending on the characteristic splicing rate $\beta$:

- **Immediate-early response** ($\beta = 5 \text{ hr}^{-1}$): Process duration = $5.0/5 = 1$ hour
- **Cell cycle progression** ($\beta = 1 \text{ hr}^{-1}$): Process duration = $5.0/1 = 5$ hours
- **Developmental transition** ($\beta = 0.1 \text{ hr}^{-1}$): Process duration = $5.0/0.1 = 50$ hours

Gene timing parameters maintain consistent biological interpretation across scales.

**Priors for other gene-specific parameters (unchanged from Study 2):**
\begin{align}
    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
\end{align}

**Stick-Breaking Cell Time Construction:**

The cell times $\{t^*_j\}_{j=1}^N$ are constructed using a stick-breaking process with fixed anchor points:

\begin{align}
t^*_1 &= 0 \quad \text{(fixed process start)} \label{eq-anchor-start} \\
t^*_N &= T_{\max} \quad \text{(fixed process end)} \label{eq-anchor-end} \\
\xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-weights} \\
t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{for } j \in \{2, \ldots, N-1\} \label{eq-stick-breaking-construction}
\end{align}

This construction automatically ensures:
- **Ordered constraint**: $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$
- **Fixed boundaries**: Process start and end are anchored
- **Efficient sampling**: $O(N)$ complexity with no constraint violations

**Cell-specific capture efficiency (unchanged from Study 2):**
\begin{align}
    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
\end{align}

@@

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$. The stick-breaking cell times provide the temporal coordinates for evaluating the piecewise analytical solutions.

@@

where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, process duration parameter $(T_{\max})$, stick-breaking cell times $(t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.

@@

\begin{align}
p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
\end{align}

@@

- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents stick-breaking cell times
- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters

This decomposition reflects the model's assumption that gene-specific kinetics, process duration, stick-breaking cell times, and technical factors contribute independently to the prior uncertainty.

@@

1. **Sample process duration**:
   \begin{align}
   {T_{\max}}^{(s)} &\sim \text{Gamma}(5.0, 1.0)
   \end{align}

>>>>

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving critical dimensional analysis inconsistencies identified in the piecewise activation framework. The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters.

@@

@@

This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations. Each pattern is defined by quantitative parameter ranges applied to **independent absolute temporal parameters** that eliminate scaling symmetry while maintaining biological interpretability.

**Independent Parameter Framework**: The classification uses independent absolute temporal parameters that are sampled directly without hierarchical scaling relationships:

\begin{align}
T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
\end{align}

This independent structure eliminates scaling symmetry by removing multiplicative relationships between temporal parameters, ensuring unique parameter identifiability while maintaining biological interpretability.

@@

- $t^*_{0,\text{on}i} < 0.0$ (absolute activation onset before observation window)
- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)

**Biological interpretation**: These genes underwent activation before the experimental observation window. During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels. The negative absolute onset time indicates that activation occurred in the "past" relative to the observation start.

@@

**Characteristics**: Genes that show complete activation-decay cycles within the observation window, exhibiting brief, pulse-like activation followed by return toward baseline.

@@

- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
- $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$ (short activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the experimental window. The constraints ensure that both the activation and subsequent decay phases are captured.

@@

**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the observation period, showing net increase over the observation window.

@@

- $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ (early activation within process duration)
- $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$ (long activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

**Biological interpretation**: These genes activate early in the observation window and maintain high expression levels throughout most of the observed period. The long activation duration relative to the observation window means that decay back to baseline is either minimal or occurs beyond the observation period.

@@

### Biological Interpretation of Timescale Prior Hyperparameters

The process duration $T_{\max}$ determines the temporal scope of observable dynamics and requires careful hyperparameter selection based on biological expectations and experimental context.

#### **Dynamical Range and Steady-State Achievement**

In the dimensionless framework where the reference rate is 1, the process duration $T_{\max}$ determines what fraction of steady-state dynamics will be observable:

$$\text{Steady-state approach} = 1 - e^{-T_{\max}}$$

**Key thresholds:**

- $T_{\max} = 1$: Observes 63% approach to steady state
- $T_{\max} = 3$: Observes 95% approach to steady state
- $T_{\max} = 5$: Observes >99% approach to steady state

#### **Hyperparameter Strategy and Biological Prior Knowledge**

**Conservative Prior (Expect Complete Dynamics):**
$$T_{\max} \sim \text{Gamma}(5.0, 1.0) \quad \text{(mean = 5, variance = 5)}$$

- **Biological assumption**: Process likely spans ~5 characteristic timescales
- **Dynamical expectation**: Will observe near-complete approach to steady state
- **Data burden**: Data must provide evidence if process is shorter than expected
- **Use case**: When biological knowledge suggests complete developmental programs

**Permissive Prior (Let Data Determine Range):**
$$T_{\max} \sim \text{Gamma}(1.0, 0.2) \quad \text{(mean = 5, variance = 25)}$$

- **Biological assumption**: Uncertain about process duration, wide range possible
- **Dynamical expectation**: Could observe early transients or complete dynamics
- **Data burden**: Data determines whether short or long timescales are needed
- **Use case**: When process duration is unknown, exploratory analysis

**Skeptical Prior (Assume Short Processes):**
$$T_{\max} \sim \text{Gamma}(1.0, 1.0) \quad \text{(mean = 1, variance = 1, heavy tail)}$$

- **Biological assumption**: Most cellular processes are rapid (early dynamics only)
- **Dynamical expectation**: Primarily observe transient phase (63% approach)
- **Data burden**: Data must strongly support longer timescales if present
- **Use case**: When expecting rapid responses, stress conditions, acute perturbations

#### **Prior Selection Guidelines**

**Choose hyperparameters based on experimental context:**

1. **Developmental processes**: Use conservative prior (Gamma(5,1)) - expect complete programs
2. **Drug responses**: Use skeptical prior (Gamma(1,1)) - expect rapid initial responses
3. **Unknown systems**: Use permissive prior (Gamma(1,0.2)) - let data decide
4. **Time-course validation**: Match prior to known experimental duration

**Sensitivity analysis recommendation:**
Test model robustness across different prior specifications to ensure conclusions are not prior-dependent.

#### **Connection to Gene Parameter Priors**

With independent parameterization, gene timing parameters should be calibrated to $T_{\max}$ expectations:

\begin{align}
t^*_{0,\text{on}i} &\sim \text{Normal}(0.3 \mathbb{E}[T_{\max}], \sigma_{on}^2) \quad \text{(expect activation at 30% of process)} \\
\delta^*_i &\sim \text{LogNormal}(\log(0.4 \mathbb{E}[T_{\max}]), \sigma_{\delta}^2) \quad \text{(expect 40% duration activation)}
\end{align}

This ensures gene parameters are scaled appropriately to the expected process duration while maintaining independence and supporting the three piecewise activation patterns (pre-activation, transient, sustained).

### Identifiability Proof for Independent Parameterization

**Theorem:** The piecewise activation likelihood with independent parameterization is not invariant under parameter scaling.

**Proof:** For any scaling factor $\lambda \neq 1$:

1. **Gene parameters are sampled independently**: $t^*_{0,\text{on}i} \not\propto T_{\max}$ and $\delta^*_i \not\propto T_{\max}$

2. **Scaling cell times but not gene parameters changes piecewise phase assignments**:
   - Original: Cell $j$ in phase determined by comparing $t^*_j$ vs. $t^*_{0,\text{on}i}$
   - Scaled: Cell $j$ in phase determined by comparing $\lambda t^*_j$ vs. $t^*_{0,\text{on}i}$ (unchanged)

3. **The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$**:
   - Phase boundaries change when cell times are scaled but gene times are not
   - Different phase assignments lead to different analytical solution forms

4. **Therefore**: $\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$ □

**Corollary:** Parameter recovery is uniquely identifiable because no two distinct parameter configurations can produce identical likelihoods.

@@

The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when observation begins. This extension is mathematically valid and biologically realistic:

@@

1. **Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$

@@

3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the observation window

@@

- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window

### Constraint Feasibility Analysis for Independent Parameterization

The pattern constraints were optimized using a systematic CDF-based analysis to ensure mathematical feasibility while preserving biological interpretability under the independent parameterization framework.

@@

# Prior hyperparameters (updated for independent absolute parameterization)
# Assuming T_max ~ Gamma(5.0, 1.0) with E[T_max] = 5.0
E_T_max = 5.0
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},                           # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 0.3 * E_T_max, 'scale': 0.8},             # Normal(1.5, 0.8²), absolute onset
    'delta_star': {'loc': np.log(0.4 * E_T_max), 'scale': 0.45},   # log(2.0), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for independent parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before observation window
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 0.3 * E_T_max),         # Early enough to complete cycle
        ('delta_star', '<', 0.4 * E_T_max),        # Short pulse duration
        ('R_on', '>', 2.0),                        # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),                    # Activation within observation window
        ('t_on_star', '<', 0.3 * E_T_max),         # Early activation onset
        ('delta_star', '>', 0.5 * E_T_max),        # Long activation duration
        ('R_on', '>', 2.0),                        # Strong fold change
    ]
}

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while using independent temporal parameterization to eliminate scaling symmetry. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of independent temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

# hyperparameters for independent time structure
pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)

@@

# independent time parameters
pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)

# latent time coordinates and cell-specific parameters
pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
pgm.add_node("lambda_j", r"$\lambda_j$", 5.5, 3)

@@

# independent time structure edges
pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
pgm.add_edge("beta_T", "T_max", plot_params=edge_params)

# independent time edges (no hierarchical scaling)
pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)

@@

The model combines independent temporal parameterization with piecewise constant transcription rates and steady-state initial conditions to eliminate scaling symmetry while enabling rich gene expression dynamics.

@@

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **independent temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- $T_{\max}$ determines the process duration and constrains cell temporal positions via $t^*_j \sim \text{Uniform}(0, T_{\max})$

**Detailed priors for independent piecewise activation parameters:**

The priors are designed to encode biologically reasonable parameter ranges while eliminating scaling symmetry through independent parameterization:

@@

**Independent temporal parameters:**
\begin{align}
    T_{\max} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \quad \text{(process duration)} \\
    t^*_j &\sim \text{Uniform}(0, T_{\max}) \quad \text{(cell temporal positions)} \\
    t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{on}, \sigma_{on}^2) \quad \text{(absolute activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(absolute activation duration)}
\end{align}

**Pattern classification framework**: The three expression patterns are defined by constraints on the independent temporal parameters:

- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$, $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$, and $R_{\text{on},i} > 2.0$

**Biological interpretation of independent prior choices:**

@@

**Independent temporal parameters:**

- **$T_{\max} \sim \text{Gamma}(\alpha_{T}, \beta_{T})$**: Process duration with hyperparameters chosen based on experimental context (conservative: Gamma(5,1), permissive: Gamma(1,0.2), skeptical: Gamma(1,1)).

- **$t^*_{0,\text{on}i} \sim \text{Normal}(\mu_{on}, \sigma_{on}^2)$**: Absolute activation onset time, typically centered around $0.3 \mathbb{E}[T_{\max}]$ to expect activation at 30% of process duration. **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\delta^*_i \sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2)$**: Absolute activation duration, typically centered around $0.4 \mathbb{E}[T_{\max}]$ to expect 40% duration activation. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): Probability depends on $\mu_{on}$ and $\sigma_{on}$ choices
- **Transient** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i < 0.4 \mathbb{E}[T_{\max}]$): Balanced probability for observable complete cycles
- **Sustained** ($0 < t^*_{0,\text{on}i} < 0.3 \mathbb{E}[T_{\max}]$ and $\delta^*_i > 0.5 \mathbb{E}[T_{\max}]$): Balanced probability for long-duration activation

**Independent parameterization benefits:**

- **Eliminates scaling symmetry**: No multiplicative relationships between temporal parameters
- **Unique identifiability**: Each parameter configuration produces a unique likelihood
- **Biological interpretability**: All parameters have direct biological meaning
- **Flexible prior specification**: Gene timing parameters can be calibrated to expected process duration

**Priors for other gene-specific parameters:**
\begin{align}
    U_{0i} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
\end{align}

**Cell-specific capture efficiency:**
\begin{align}
    \lambda_j &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
\end{align}

@@

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$. All temporal parameters are now independent, eliminating scaling symmetry.

@@

where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, independent time parameters $(T_{\max}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.

@@

\begin{align}
p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-independent-integration}
\end{align}

@@

- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters

This decomposition reflects the model's assumption that gene-specific kinetics, cell-specific temporal positions, and technical factors contribute independently to the prior uncertainty, with no hierarchical scaling relationships.

@@

1. **Sample process duration**:
   \begin{align}
   {T_{\max}}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T})
   \end{align}
